Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-06-14 manuscript.qmd
mention llh as prior in appendix

In @tbl-data-unknowns we show the combinations of the midline extension status and state of the contralateral LNL II. Using only available information, we would get to 6 out of 29<!-- TODO: compute dynamically --> patients for the prevalence of contralateral LNL II invovlement among the early T-category patients with midline extension. But the column with the "unknown" numbers suggests that the true prevalence could range from 6 out of 29 + 59 (7%)<!-- TODO: compute dynamically --> up to 12 out of 29 + 6 (34%)<!-- TODO: compute dynamically -->. Assuming that every one of the possible combinations has the same a priori probability (uniform prior) and accouting for permutations via the binomial factor, the posterior marginalized over the unknowns could be written as

$$
\begin{aligned}
p(x; \alpha=1, \beta=1) &= \sum_{m=0}^{65} \sum_{\ell=0}^{\min(m,6)} \binom{n+m}{k+\ell} x^{k+\ell} (1-x)^{n+m - (k+\ell)} \cdot \frac{1}{\operatorname{B}(\alpha, \beta)} x^\alpha (1-x)^\beta \\
&= \sum_{m=0}^{65} \sum_{\ell=0}^{\min(m,6)} \binom{n+m}{k+\ell} x^{k+\ell+1} (1-x)^{n+m - (k+\ell) + 1}
\end{aligned}
$$

@@

The distribution that marginalizes over the unknown midline extension cases is shifted towards a lower prevalence. This makes intuitive sense: With only six patients among the 65 with unknown midline extension status, the upper bound for the prevalence has moved down compared to six out of 29.

>>>>

In @tbl-data-unknowns we show the combinations of the midline extension status and state of the contralateral LNL II. Using only available information, we would get to 6 out of 29<!-- TODO: compute dynamically --> patients for the prevalence of contralateral LNL II invovlement among the early T-category patients with midline extension. But the column with the "unknown" numbers suggests that the true prevalence could range from 6 out of 29 + 59 (7%)<!-- TODO: compute dynamically --> up to 12 out of 29 + 6 (34%)<!-- TODO: compute dynamically -->. We assume that every one of the possible combinations has the same a priori probability (uniform prior) but we do need to account for permutations via the binomial factor. In general, let us say we have $n$ patients with midline extension, of which $k$ harbor metastases in the contralateral LNL II. And we have $m$ patients with unknown midline extension, of which $\ell$ presented with diasease in contralateral LNL II, then the posterior distribution - marginalized over the unknown data - over the true prevalence $x \in \left[ 0, 1 \right]$ is given by:

$$
p(x \mid n, k, m, \ell, \alpha, \beta) = \left[ \frac{1}{m} \sum_{i=0}^m \frac{1}{\min(i,\ell)} \sum_{j=0}^{\min(i,\ell)} \frac{P(i, j \mid x )}{\binom{n}{k}} \right] \cdot P(n, k \mid x) \cdot \frac{1}{\operatorname{B}(\alpha, \beta)} x^\alpha (1-x)^\beta
$$

Note the two binomial likelihoods: one being marginalized over and one acting implicitly as a prior to the summed terms. This role of the likelihood over the data with known midline extension as a prior to the part where that data is missing corrects for the fact that there are highly unlikely combinations among those we marginalize over. E.g., expecting that exactly those 6 patients with contralateral LNL II involvemnt are also coincidentially the only ones of the 65 with a tumor crossing the mid-sagittal line is very unlikely. This is accounted for by the fact that this possibility has only one permutation and it is weakened by the likelihood over the data with known midline extension. With a uniform prior beta distribution ($\alpha=1$ and $\beta=1$), the above simplifies to:

$$
p(x \mid n, k, m, \ell) = \left[ \frac{1}{m} \sum_{i=0}^m \frac{1}{\min(i,\ell)} \sum_{j=0}^{\min(i,\ell)} \binom{n+i}{k+j} x^j (1 - x)^{i-j} \right] \cdot x^{k+1} (1 - x)^{n-k+1}
$$

@@

The distribution that marginalizes over the unknown midline extension cases is shifted towards a lower prevalence. This makes intuitive sense: With only 6 patients among the 65 with unknown midline extension status, the upper bound for the prevalence has moved down compared to 6 out of 29.
